---
title: "ESM 244 Lab 1"
author: "Logan Ossentjuk"
date: "1/10/2022"
output: html_document
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE} 
#Includes what packages are libraries
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(sf)
library(tmap)
### Update packages(ask = FALSE)
```


### Read in the data 

```{r}
sf_trees <- read_csv(here('data', 'sf_trees', 'sf_trees.csv'))
```


### Part 1: Wrangling and ggplot review 

**Example 1:** Find couunts of observation by `legal_status` & wrangle a bit.

```{r}
### Method 1: group_by() %>%  summarize()

sf_trees %>% 
  group_by(legal_status) %>% #groups sf_trees by legal_status
  summarize(tree_count = n()) #outputs the count of each value in legal status

### Method 2: different way plus a few new functions

top_5_status <- sf_trees %>% 
  count(legal_status) %>% #outputs count of each value and groups by legal_status
  drop_na(legal_status) %>%  #drops any na values in legal_status
  rename(tree_count = n) %>% 
  relocate(tree_count) %>%  #brings given column name to the front of the data set
  slice_max(tree_count, n = 5) %>%  #slices out the top 5 values in tree_count
  arrange(-tree_count) #arranges by tree_count descending (b/c of - sign) (can also do arrange(desc(tree_count))

```


Make a graph of the top 5 from above: 

```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), y = tree_count)) +
  geom_col(fill = 'darkgreen') +
  labs(x = 'Legal status', y = 'Tree count') +
  coord_flip() + #switches columns to face horizontally 
  theme_minimal()

```


### **Example 2:** Only going to keep bservations where legal status is "Permitted Site" and caretaker is "MTA", and stores is `permitted_data_df`

shift-command-c to comment/uncomment a line quickly
```{r}
# sf_trees$legal_status %>% unique()
# unique(sf_trees$caretaker)
permitted_data_df <- sf_trees %>% 
  filter(legal_status %in% c('Permitted Site', 'Private'), caretaker == 'MTA') #, or & for and, | for or=
```

### **Example 3:** Only keep Blackwood Acacia trees, and then only keep columns `legal_status`, `date`, `latitude`, and stores as `blackwood_acacia_df`

```{r}
blackwood_acacia_df <- sf_trees %>% 
  filter(str_detect(species, 'Blackwood Acacia')) %>%  #str_detect finds a given string within a row even if there is other text within it 
  select(legal_status, date, lat = latitude, lon = longitude)

### Make a little graph of locations. 
ggplot(data = blackwood_acacia_df, aes( x = lon, y = lat)) +
  geom_point(color = 'darkgreen')

```


### **Example 4:** use `tidyr::separate()` 

```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c('spp_scientific', 'spp_common'), sep = ' :: ')
```

### **Example 5:**  use `tidyr::unite()`

```{r}
ex_5 <- sf_trees %>% 
  unite('id_status', tree_id, legal_status, sep = '_COOL_')
```




